---
title: "The 'symbolicQspray' package"
author: "Stéphane Laurent"
date: "`r Sys.Date()`"
output: github_document
editor_options: 
  chunk_output_type: console
---

***Multivariate polynomials with symbolic coefficients.*** 

<!-- badges: start -->
[![R-CMD-check](https://github.com/stla/symbolicQspray/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/stla/symbolicQspray/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, message = FALSE)
```

___

These notes about the **symbolicQspray** package assume that the reader is a 
bit familiar with the [**qspray** package](https://github.com/stla/qspray) 
and the [**ratioOfQsprays** package](https://github.com/stla/ratioOfQsprays).

A `symbolicQspray` object represents a multivariate polynomial whose 
coefficients are fractions of polynomials with rational coefficients. 
Actually (see our discussion in the next section), a `symbolicQspray` object 
represents a *multivariate polynomial with parameters*. The parameters are 
the variables of the fractions of polynomials.

To construct a `symbolicQspray` polynomial, use `qlone` (from the **qspray** 
package) to introduce the parameters and use `Qlone` to introduce the variables 
of the polynomial:

```{r}
library(symbolicQspray)
f <- function(a1, a2, X1, X2, X3) {
  (a1/(a2^2+1)) * X1^2*X2  +  (a2+1) * X3  +  a1/a2
}
# parameters, the variables occurring in the coefficients:
a1 <- qlone(1)
a2 <- qlone(2)
# variables:
X1 <- Qlone(1)
X2 <- Qlone(2)
X3 <- Qlone(3)
# the 'symbolicQspray':
( Qspray <- f(a1, a2, X1, X2, X3) )
```

The fractions of polynomials such as the first coefficient `a1/(a2^2+1)` 
in the above example are 
[**ratioOfQsprays**](https://github.com/stla/ratioOfQsprays) objects, 
and the numerator and the denominator of a `ratioOfQsprays` are 
[**qspray**](https://github.com/stla/qspray) objects.

Arithmetic on `symbolicQspray` objects is available:

```{r}
Qspray^2
Qspray - Qspray
(Qspray - 1)^2
Qspray^2 - 2*Qspray + 1
```


## Evaluating a `symbolicQspray`

Substituting the "exterior" variables (the variables occurring in the ratios of 
polynomials, also called the *parameters* - see below) yields a `qspray` object:

```{r}
a <- c(2, "3/2")
( qspray <- evalSymbolicQspray(Qspray, a = a) )
```

Substituting the "main" variables yields a `ratioOfQsprays` object:

```{r}
X <- c(4, 3, "2/5")
( ratioOfQsprays <- evalSymbolicQspray(Qspray, X = X) )
```

There is a discutable point here. A `symbolicQspray` object represents a 
polynomial with `ratioOfQsprays` coefficients. So one can consider 
that the polynomial variables `X`, `Y` and `Z` represent some indeterminate
`ratioOfQsprays` fractions, and that it should be possible to replace them 
with `ratioOfQsprays` objects. However this is not allowed.
We will discuss that, just after checking the consistency:

```{r}
evalSymbolicQspray(Qspray, a = a, X = X)
evalQspray(qspray, X)
evalRatioOfQsprays(ratioOfQsprays, a)
a <- gmp::as.bigq(a); X <- gmp::as.bigq(X)
f(a[1], a[2], X[1], X[2], X[3])
```

Now let's turn to our promised discussion. Why is replacing the values of the 
polynomial variables with some `ratioOfQsprays` objects not allowed?

Actually my motivation to do this package was inspired by the 
**Jack polynomials**. In the context of Jack polynomials, the variables 
`X`, `Y` and `Z` represent indeterminate *numbers*, and the coefficients 
are *numbers depending on a parameter* (the Jack parameter), and it turns out 
that they are fractions of polynomials of this parameter. 
So I consider that a `symbolicQspray` is *not* a polynomial on the field 
of fractions of polynomials: I consider it is a polynomial with 
*rational coefficients depending on some parameters*. 

By the way, I'm wondering whether I should rename `symbolicQspray` to 
`parametricQspray`. Comments?..

Also note that evaluating the `ratioOfQsprays` object 
`evalSymbolicQspray(Qspray, X = X)` at `a` would make no sense if we took 
some `ratioOfQsprays` objects for the values of `X`.


## Querying a `symbolicQspray` 

The package provides some functions to perform elementary queries on a 
`symbolicQspray`:

```{r, collapse=TRUE}
numberOfVariables(Qspray)
numberOfParameters(Qspray)
numberOfTerms(Qspray)
getCoefficient(Qspray, c(2, 1))
getConstantTerm(Qspray)
isUnivariate(Qspray)
isConstant(Qspray)
```

## Transforming a `symbolicQspray`

You can derivate a `symbolicQspray` polynomial:

```{r}
derivSymbolicQspray(Qspray, 2) # derivative w.r.t. Y
```

You can permute its variables:

```{r}
swapVariables(Qspray, 2, 3) == f(a1, a2, X1, X3, X2)
```

You can perform polynomial transformations of its variables:

```{r}
changeVariables(Qspray, list(X1+1, X2^2, X1+X2+X3)) == 
  f(a1, a2, X1+1, X2^2, X1+X2+X3)
```

You can also perform polynomial transformations of its parameters:

```{r}
changeParameters(Qspray, list(a1^2, a2^2)) == f(a1^2, a2^2, X1, X2, X3)
```


## Showing a `symbolicQspray` 

You can change the way a `symbolicQspray` is printed by using 
`showSymbolicQsprayOption`:

```{r, collapse=TRUE}
showSymbolicQsprayOption(Qspray, "a") <- "x"
showSymbolicQsprayOption(Qspray, "showMonomial") <- 
  showMonomialXYZ(c("A", "B", "C"))
showSymbolicQsprayOption(Qspray, "quotientBar") <- " / "
Qspray
```

When this is possible, the result of an arithmetic operation between two 
`symbolicQspray` objects inherits the show options of the first operand:

```{r, collapse=TRUE}
set.seed(421)
( Q <- rSymbolicQspray() ) # a random symbolicQspray
Qspray + Q
```

This behavior is the same as the ones implemented in **qspray** and in 
**ratioOfQsprays**. You should be familiar with these two packages in order 
to use **symbolicQspray**. 


## Application: Jacobi polynomials

The [Jacobi polynomials](https://en.wikipedia.org/wiki/Jacobi_polynomials) 
are univariate polynomials depending on two parameters that we will denote by
`alpha` and `beta`. They are implemented in this package:

```{r}
JP <- JacobiPolynomial(2)
isUnivariate(JP)
numberOfParameters(JP)
showSymbolicQsprayOption(JP, "showRatioOfQsprays") <-
  showRatioOfQspraysXYZ(c("alpha", "beta"))
JP
```

The implementation constructs these polynomials by using the 
[recurrence relation](https://en.wikipedia.org/wiki/Jacobi_polynomials#Recurrence_relations).
This is a child game, one just has to copy the first two terms and this 
recurrence relation:

```{r, eval=FALSE}
JacobiPolynomial <- function(n) {
  stopifnot(isPositiveInteger(n))
  if(n == 0) {
    Qone()
  } else if(n == 1) {
    alpha <- qlone(1)
    beta  <- qlone(2)
    X     <- Qlone(1)
    (alpha + 1) + (alpha + beta + 2) * (X - 1)/2
  } else {
    alpha <- qlone(1)
    beta  <- qlone(2)
    X     <- Qlone(1)
    a <- n + alpha
    b <- n + beta
    c <- a + b
    K <- 2 * n * (c - n) * (c - 2)
    lambda1 <- ((c - 1) * (c * (c - 2) * X + (a - b) * (c - 2*n))) / K
    lambda2 <- (2 * (a - 1) * (b - 1) * c) / K
    (lambda1 * JacobiPolynomial(n - 1) - lambda2 * JacobiPolynomial(n - 2))
  }
}
```

Up to a factor, the 
[Gegenbauer polynomials](https://en.wikipedia.org/wiki/Gegenbauer_polynomials)
with parameter `alpha` coincide with the Jacobi polynomials with parameters 
`alpha - 1/2` and `alpha - 1/2`. Let's derive them from the Jacobi polynomials,
as an exercise. The factor can be implemented as follows (see Wikipedia for its 
formula):

```{r}
risingFactorial <- function(theta, n) {
  toMultiply <- c(theta, lapply(seq_len(n-1), function(i) theta + i))
  Reduce(`*`, toMultiply)
}
theFactor <- function(alpha, n) {
  risingFactorial(2*alpha, n) / risingFactorial((2*alpha + 1)/2, n)
}
```

Now let's apply the formula given in the Wikipedia article:

```{r}
GegenbauerPolynomial <- function(n) {
  alpha <- qlone(1)
  P <- changeParameters(
    JacobiPolynomial(n), list(alpha - "1/2", alpha - "1/2")
  )
  theFactor(alpha, n) * P
}
```

Let's check the recurrence relation given in the Wikipedia article is fulfilled:

```{r}
n <- 5
alpha <- qlone(1)
X <- Qlone(1)
(n+1)*GegenbauerPolynomial(n+1) == 
  2*(n+alpha)*X*GegenbauerPolynomial(n) - 
    (n+2*alpha-1)*GegenbauerPolynomial(n-1)
```


## Application to Jack polynomials

The **symbolicQspray** package is used by the 
[**jack**](https://github.com/stla/jackR) package to compute the Jack 
polynomials. The Jack polynomials are exactly the polynomials 
represented by the `symbolicQspray` objects: their coefficients are fractions 
of polynomials by definition, of one variable: the Jack parameter.
