---
title: "The 'symbolicQspray' package"
author: "Stéphane Laurent"
date: "`r Sys.Date()`"
output: github_document
---

*Symbolic multivariate polynomials.* 

<!-- badges: start -->
[![R-CMD-check](https://github.com/stla/symbolicQspray/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/stla/symbolicQspray/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

___

A `symbolicQspray` object represents a multivariate polynomial whose 
coefficients are fractions of polynomials. For example:

```{r, message=FALSE, collapse=TRUE}
library(symbolicQspray)
f <- function(a1, a2, X1, X2, X3) {
  (a1/(a2^2+1)) * X1^2*X2  +  (a2+1) * X3  +  a1/a2
}
# "exterior" variables, occuring in the coefficients:
a1 <- qlone(1)
a2 <- qlone(2)
# "main" variables:
X1 <- Qlone(1)
X2 <- Qlone(2)
X3 <- Qlone(3)
# the 'symbolicQspray':
( Qspray <- f(a1, a2, X1, X2, X3) )
```

The fractions of polynomials such as the first coefficient `a1/(a2^2+1)` 
in the above example are 
[**ratioOfSprays**](https://github.com/stla/ratioOfQsprays) objects, 
and the numerator and the denominator of a `ratioOfQsprays` are 
[**qspray**](https://github.com/stla/qspray) objects.

Arithmetic on `symbolicQspray` objects is available:

```{r, message=FALSE, collapse=TRUE}
Qspray^2
Qspray - Qspray
(Qspray - 1)^2
Qspray^2 - 2*Qspray + 1
```


## Evaluating a `symbolicQspray`

Substituting the "exterior" variables (the variables occurring in the ratios of 
polynomials) yields another `symbolicQspray`:

```{r, message=FALSE, collapse=TRUE}
a <- c(2, "3/2")
( Q <- evalSymbolicQspray(Qspray, a = a) )
```

Substituting the "main" variables yields a `ratioOfQsprays` object:

```{r, message=FALSE, collapse=TRUE}
X <- gmp::as.bigq(c(4, 3, "2/5"))
( ratioOfQsprays <- evalSymbolicQspray(Qspray, X = X) )
```

Checking the consistency:

```{r, message=FALSE, collapse=TRUE}
evalSymbolicQspray(Qspray, a = a, X = X)
evalSymbolicQspray(Q, X = X)
evalRatioOfQsprays(ratioOfQsprays, a)
f(gmp::as.bigq(a[1]), gmp::as.bigq(a[2]), X[1], X[2], X[3])
```

Actually we can more generally use some `ratioOfQsprays` objects for the 
values of `X`, since a `symbolicQspray` object is a polynomial with 
`ratioOfQsprays` coefficients:

```{r, message=FALSE, collapse=TRUE}
X <- list(
  qlone(1)/(1+qlone(1)), qlone(2)/qlone(1)^2, (qlone(2) + qlone(3))/qlone(1)
)
evalSymbolicQspray(Qspray, a = a, X = X)
evalSymbolicQspray(Q, X = X)
```


## Querying a `symbolicQspray` 

The package provides some functions to perform elementary queries on a 
`symbolicQspray`:

```{r, collapse=TRUE}
numberOfVariables(Qspray)
numberOfTerms(Qspray)
getCoefficient(Qspray, c(2, 1))
getConstantTerm(Qspray)
isUnivariate(Qspray)
isConstant(Qspray)
```


## Showing a `symbolicQspray` 

You can change the way a `symbolicQspray` is printed by using 
`showSymbolicQsprayOption`:

```{r, collapse=TRUE}
showSymbolicQsprayOption(Qspray, "a") <- "x"
showSymbolicQsprayOption(Qspray, "X") <- "A"
showSymbolicQsprayOption(Qspray, "quotientBar") <- " / "
Qspray
```

