---
title: "The 'symbolicQspray' package"
author: "Stéphane Laurent"
date: "`r Sys.Date()`"
output: github_document
---

*Symbolic multivariate polynomials.* 

<!-- badges: start -->
[![R-CMD-check](https://github.com/stla/symbolicQspray/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/stla/symbolicQspray/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

___

A `symbolicQspray` object represents a multivariate polynomial whose 
coefficients are fractions of polynomials. For example:

```{r, message=FALSE, collapse=TRUE}
library(symbolicQspray)
f <- function(a1, a2, X1, X2, X3) {
  (a1/(a2^2+1)) * X1^2*X2  +  (a2+1) * X3
}
# "inner" variables:
a1 <- qlone(1)
a2 <- qlone(2)
# "main" variables:
X1 <- Qlone(1)
X2 <- Qlone(2)
X3 <- Qlone(3)
# the 'symbolicQspray':
( Qspray <- f(a1, a2, X1, X2, X3) )
```

Arithmetic on `symbolicQspray` objects is available:

```{r, message=FALSE, collapse=TRUE}
Qspray^2
Qspray - Qspray
```

Substituting the "inner" variables (the variables occurring in the ratios of 
polynomials) yields a `qspray` object:

```{r, message=FALSE, collapse=TRUE}
a <- c(2, "3/2")
( qspray <- evalSymbolicQspray(Qspray, a = a) )
```

Substituting the "main" variables yields a `ratioOfQsprays` object:

```{r, message=FALSE, collapse=TRUE}
X <-  c(4, 3, "2/5")
( ratioOfQsprays <- evalSymbolicQspray(Qspray, X = X) )
```

Checking the consistency:

```{r, message=FALSE, collapse=TRUE}
evalSymbolicQspray(Qspray, a = a, X = X)
evalQspray(qspray, X)
evalRatioOfQsprays(ratioOfQsprays, a)
```

